Forecasting the United States 2024 Presidential Election

### 2024-11-04 paper/paper.qmd
syntax edits

Data was analyzed through the R programming software [@citeR] and packages such as tidyverse [@tidyverse], ggplot2 [@ggplot2], knitr [@knitr], and dplyr [@dplyr] were used to help download, clean, simulate, analyze, and test the data. For model construction and analysis, the following packages were used: rstanarm [@rstanarm1, @rstanarm2], brms [@brms1, @brms2, @brms3], and bayesplot [@bayesplot1, @bayesplot2]

@@

All model construction and analysis was conducted using R (@citeR) with the following key packages:

- **tidyverse:** For data manipulation and visualization. (@tidyverse)
- **rstanarm:** For Bayesian model fitting using Stan. (@rstanarm1, @rstanarm2)
- **brms:** Alternative package for Bayesian regression modeling. (@brms1, @brms2, @brms3)
- **bayesplot:** For extensive model diagnostics and visualization.(@bayesplot1, @bayesplot2)

@@

For states lacking recent polling data, we incorporated historical averages of Democratic and Republican support from elections since 2000. This data was collected from a GitHub repository (@hist_state) which collected their data from Wikipedia. This decision is based on the assumption that certain states are unlikely to change their voting patterns, reducing the necessity for current polling data. The ten states that do not have polling data have not changed parties in at least the last 20-plus years (@sec-voting-hist). Alternatively, imputation methods or political regional averages could have been used, but historical averages are used for their simplicity and relevance.

@@

Ideally, multiple models would be created. Then, model selection techniques would be used to determine the best model for predicting the outcome of the US election. Due to time constraints, the following models were considered but not implemented. A set of Bayesian models with state as a random effect and models excluding specific predictors like Transparency Score. More complex models with additional interaction terms or non-linear components were also considered

@@

@table-2 displays each candidate's chance of winning and the average number of Elector Votes they received over the 1000 simulated elections. Our model predicts that Harris has a 67.4% chance of winning the US presidential election. While Trump only has a 31.7% chance of winning. There is also a 0.9% chance of a tie On average across the 1,000 simulations Harris wins 279.33 Electoral Votes, while Trump wins on average 258.67 votes.

```{r}
#| label: table-2
#| fig-cap: Election Prediction Results
#| echo: false
#| warning: false
#| message: false

@@

Prediction betting markets provide a new data source that may be valuable for predicting the outcome of the US election. These online betting markets allow people to buy shares that represent an outcome of some event. If the event occurs then the share is worth $1, if not then the share's price falls to 0. This means the live share price reflects the probability of the event occurring. For example, for the outcome of the presidential election market, if Trump's share price is 65 cents this means the market is predicting there is a 65% chance that Trump will win the election. Recently the trading volume on these markets is significant, which indicates that these markets may be helpful in predicting election outcomes. Some papers have included data from these markets in their models to improve their prediction accuracy (@market_paper).

>>>>

Data was analyzed through the R programming software [@citeR] and packages such as `tidyverse` [@tidyverse], `ggplot2` [@ggplot2], `knitr` [@knitr], and `dplyr` [@dplyr] were used to help download, clean, simulate, analyze, and test the data. For model construction and analysis, the following packages were used: `rstanarm` [@rstanarm1, @rstanarm2], `brms` [@brms1, @brms2, @brms3], and `bayesplot` [@bayesplot1, @bayesplot2].

@@

All model construction and analysis was conducted using R [@citeR] with the following key packages:

- **tidyverse:** For data manipulation and visualization [@tidyverse].
- **rstanarm:** For Bayesian model fitting using Stan [@rstanarm1; @rstanarm2].
- **brms:** Alternative package for Bayesian regression modeling [@brms1; @brms2; @brms3].
- **bayesplot:** For extensive model diagnostics and visualization [@bayesplot1; @bayesplot2].

@@

For states lacking recent polling data, we incorporated historical averages of Democratic and Republican support from elections since 2000. This data was collected from a GitHub repository [@hist_state] which collected their data from Wikipedia. This decision is based on the assumption that certain states are unlikely to change their voting patterns, reducing the necessity for current polling data. The ten states that do not have polling data have not changed parties in at least the last 20-plus years (@sec-voting-hist). Alternatively, imputation methods or political regional averages could have been used, but historical averages are used for their simplicity and relevance.

@@

Ideally, multiple models would be created. Then, model selection techniques would be used to determine the best model for predicting the outcome of the US election. Due to time constraints, the following models were considered but not implemented. A set of Bayesian models with state as a random effect and models excluding specific predictors like Transparency Score. More complex models with additional interaction terms or non-linear components were also considered.

@@

@fig-predic displays each candidate's chance of winning and the average number of Elector Votes they received over the 1000 simulated elections. Our model predicts that Harris has a 67.4% chance of winning the US presidential election. While Trump only has a 31.7% chance of winning. There is also a 0.9% chance of a tie On average across the 1,000 simulations Harris wins 279.33 Electoral Votes, while Trump wins on average 258.67 votes.

```{r}
#| label: fig-predic
#| fig-cap: Election Prediction Results
#| echo: false
#| warning: false
#| message: false

@@

Prediction betting markets provide a new data source that may be valuable for predicting the outcome of the US election. These online betting markets allow people to buy shares that represent an outcome of some event. If the event occurs then the share is worth $1, if not then the share's price falls to 0. This means the live share price reflects the probability of the event occurring. For example, for the outcome of the presidential election market, if Trump's share price is 65 cents this means the market is predicting there is a 65% chance that Trump will win the election. Recently the trading volume on these markets is significant, which indicates that these markets may be helpful in predicting election outcomes. Some papers have included data from these markets in their models to improve their prediction accuracy [@market_paper].
